---
title: 'Practical Machine Learning: Prediction Assignment'
output: html_document
---
```{r}
training_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testing_url  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

download.file(training_url, "pml-training.csv", mode = "wb")
download.file(testing_url, "pml-testing.csv", mode = "wb")

training <- read.csv("pml-training.csv", na.strings = c("", "NA"))
testing <- read.csv("pml-testing.csv", na.strings = c("", "NA"))

dim(training)
dim(testing)
```
```{r}
library(caret)
library(randomForest)

# Remove columns with mostly missing values
training <- training[, colSums(is.na(training)) < 0.9 * nrow(training)]

testing <- testing[, names(training)[names(training) != "classe"]]

# Remove identification/time-window variables
remove_cols <- c(
  "X",
  "user_name",
  "raw_timestamp_part_1",
  "raw_timestamp_part_2",
  "cvtd_timestamp",
  "new_window",
  "num_window"
)

remove_cols <- intersect(remove_cols, names(training))

training <- training[, !names(training) %in% remove_cols]
testing <- testing[, !names(testing) %in% remove_cols]
```

```{r}
set.seed(123)

inTrain <- createDataPartition(
  training$classe,
  p = 0.75,
  list = FALSE
)

trainData <- training[inTrain, ]
validData <- training[-inTrain, ]

dim(trainData)
dim(validData)
```
 
```{r}
# Make sure the outcome is a factor
trainData$classe <- as.factor(trainData$classe)

# Remove rows containing missing values
trainData <- na.omit(trainData)
validData <- na.omit(validData)

# Train Random Forest
model <- randomForest(
  classe ~ .,
  data = trainData,
  ntree = 100
)

model
```
 
```{r}
# Make predictions on the validation data
validation_prediction <- predict(model, newdata = validData)

# Make sure both are factors with the same levels
validation_prediction <- factor(
  validation_prediction,
  levels = c("A", "B", "C", "D", "E")
)

validData$classe <- factor(
  validData$classe,
  levels = c("A", "B", "C", "D", "E")
)

confusionMatrix(
  data = validation_prediction,
  reference = validData$classe
)
```
 
```{r}
# Predict the 20 test cases
test_prediction <- predict(model, newdata = testing)

# Display the predictions
test_prediction
```

## Results

The Random Forest model was evaluated using the validation dataset. The
validation accuracy was calculated and used to estimate the out-of-sample
error. The trained model was then applied to the 20 observations in the
test dataset.

The predictions for the 20 test cases were generated using the final model.

```{r}
# Create validation predictions
validation_prediction <- predict(model, newdata = validData)

# Make sure the factors have the same levels
validation_prediction <- factor(
  validation_prediction,
  levels = c("A", "B", "C", "D", "E")
)

validData$classe <- factor(
  validData$classe,
  levels = c("A", "B", "C", "D", "E")
)

# Calculate accuracy and out-of-sample error
accuracy <- mean(validation_prediction == validData$classe)
out_of_sample_error <- 1 - accuracy

accuracy
out_of_sample_error
```
## Introduction

The goal of this project is to predict the manner in which participants
performed barbell lifts using accelerometer measurements. The response
variable is the classe variable, which contains five exercise categories.

## Data Preparation

The training data were divided into training and validation sets. Variables
with large amounts of missing data and identification-related variables were
removed before modeling.

## Model

A Random Forest classification model was used to predict the classe variable.
The model was trained using the training portion of the dataset.

## Cross Validation

The training data were divided into a training set and a validation set.
The validation set was used to estimate how well the model performs on
previously unseen observations.

## Results

The Random Forest model achieved a validation accuracy of 99.59%.
The estimated out-of-sample error was approximately 0.41%.

## Test Set Predictions

The final model was applied to the 20 observations in the test dataset.
The resulting predictions were submitted to the Course Project Prediction
Quiz.
 
 